---
title: "CreatePhyloSeqObjects"
author: "Catherine Purse"
date: "2024-10-15"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* Load libraries
```{r}
library(phyloseq)
library(tidyverse)
library(microbiome)
library(microViz)
```

* Function to create Phyloseq object from Metaphlan profile (https://rdrr.io/github/g-antonello/gautils2/src/R/metaphlan_to_phyloseq.R)
```{r}
metaphlan_to_phyloseq <- function(mpa,
                                  metadata = NULL,
                                  version = 4,
                                  verbose = TRUE,
                                  tax_lvl = "Species"){


  if(version == 4){
    if(is.character(mpa)){
      # load raw metaphlan data
      mpa <- data.table::fread(mpa) %>%
        as.data.frame()
    }
  }
    if(version == 3){
      if(is.character(mpa)){
        # load raw metaphlan data
        mpa <- data.table::fread(mpa,skip = 1) %>%
          as.data.frame()
      }
    }
  # find for each row, to which depth of taxonomy it arrives (as integers)
  tax_lengths <- mpa$clade_name %>%
    strsplit("|", fixed = T) %>%
    sapply(length)
  # remove first element,we want to keep it
  tax_lengths <- tax_lengths[-1]
  # get integer equivalent of the taxonomic level we want
  tax_lvl_int <- match(tax_lvl, c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "SGB"))
  # subset the otu table, so to keep only UNASSIGNED on top and the rows exactly to te taxonomic level
  otu_cleaned <- mpa[c(T, tax_lengths == tax_lvl_int), 2:ncol(mpa)]

  if(!is.null(metadata)){
    inters_names <- intersect(colnames(otu_cleaned), rownames(metadata))
    if(verbose){
      if(length(colnames(otu_cleaned)[!(colnames(otu_cleaned) %in% inters_names)])!= 0){
        cat("Metaphlan table samples lost: ")
        cat(colnames(otu_cleaned)[!(colnames(otu_cleaned) %in% inters_names)], sep = " ")
        cat("\n")
        cat("\n")
      }

      if(length(rownames(metadata)[!(rownames(metadata) %in% inters_names)]) != 0){
        cat("Metadata table samples lost: ")
        cat(rownames(metadata)[!(rownames(metadata) %in% inters_names)], sep = " ")
        cat("\n")
        cat("\n")
      }
    }

    otu_cleaned <- otu_cleaned[, inters_names]
    metadata_cleaned <- metadata[match(inters_names, rownames(metadata)),]

  }

  # create taxonomy table, filling empty columns with NA

  taxonomy_tab <- mpa[c(T, tax_lengths == tax_lvl_int), 1] %>%
    as.data.frame() %>%
    set_names("clade_name") %>%
    separate(col="clade_name", into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "SGB")[1:tax_lvl_int],sep = "\\|", fill = "right") %>%
    as.matrix()
  # rename the UNKNOWN clade as all unknown, otherwise we can't assign rownames
  taxonomy_tab[1,] <- rep("UNKNOWN",tax_lvl_int)

  #assign rownames to the tax level chosen
  rownames(taxonomy_tab) <- paste0("otu", seq_len(nrow(otu_cleaned)))
  rownames(otu_cleaned) <- rownames(taxonomy_tab)
  


    profiles <- phyloseq(otu_table(as.matrix(otu_cleaned), taxa_are_rows = TRUE),
                         tax_table(taxonomy_tab),
                         sample_data(metadata_cleaned, errorIfNULL = FALSE)
    )

    return(profiles)
}
```


* Load data
```{r}
#load metadata and reassign column 1 as rownames
metadata <- read_csv("Metadata_6189_Oct24.csv") %>% 
  remove_rownames %>% 
  column_to_rownames(var =  "sample")

mpa_withunknown <- read_tsv("6189_merged_ab.txt", skip=1)
```

## Colour palette
```{r}
cb_palette <- c("#ee8866", "#77aadd","#d3d3d3","#FDF28E","#8ED29B","#bbccee", "#AFC30A","#cc99cc","#C1E49C","#dd7788", "#009E73")
```

## Kingdom level

* Parse metaphlan data into phyloseq object
```{r}
kingdom_physeq <- metaphlan_to_phyloseq(
  mpa_withunknown,
  metadata = metadata,
  version = 4,
  verbose = TRUE,
  tax_lvl = "Kingdom"
)
```

```{r}
kingdom_physeq <- kingdom_physeq %>% tax_fix()
```

* Remove the leading characters (e.g. p__)
```{r}
# view part of the tax table
phyloseq::tax_table(kingdom_physeq)

# replace _[A-Z] with ""
tax_table(kingdom_physeq)[, colnames(tax_table(kingdom_physeq))] <- gsub(tax_table(kingdom_physeq)[, colnames(tax_table(kingdom_physeq))],     pattern = "[a-z]__", replacement = "")

# check tax_table again
phyloseq::tax_table(kingdom_physeq)
```

* Remove unneeded samples
```{r}
# Filter kingdom level phyloseq object to remove Eukaryotes
kingdom_physeq <- subset_taxa(kingdom_physeq, Kingdom!="Eukaryota")

# Prune "empty" samples
kingdom_physeq <- prune_samples(sample_sums(kingdom_physeq) > 0, kingdom_physeq)

# Remove faecal samples
kingdom_physeq <- ps_filter(kingdom_physeq, type != "Faecal")
```

* Test plot
```{r}
plot_bar(kingdom_physeq, fill="Kingdom")
```

* Identify number of samples containing archaea
```{r}
# Identify taxa that are classified as archaea
archaea_taxa <- taxa_names(subset_taxa(kingdom_physeq, Kingdom == "Archaea"))

# Subset the phyloseq object to include only archaea
archaea_physeq <- prune_taxa(archaea_taxa, kingdom_physeq)

# Check which samples have non-zero abundance for archaea
archaea_present_samples <- sample_sums(archaea_physeq) > 0

# Count the number of samples with archaea
num_samples_with_archaea <- sum(archaea_present_samples)

# Output the result
cat("Number of samples with archaea:", num_samples_with_archaea, "\n")
```

* Export phyloseq object for use in other scripts
```{r}
saveRDS(kingdom_physeq, file = "kingdom_phyloseq.rds")
```


## Phylum level

* Parse metaphlan data into phyloseq object
```{r}
phylum_physeq <- metaphlan_to_phyloseq(
  mpa_withunknown,
  metadata = metadata,
  version = 4,
  verbose = TRUE,
  tax_lvl = "Phylum"
)
```

* Fill in any empty cells using microviz function
```{r}
phylum_physeq <- phylum_physeq %>% tax_fix()
```

* Remove the leading characters (e.g. p__)
```{r}
# view part of the tax table
phyloseq::tax_table(phylum_physeq)

# replace _[A-Z] with ""
tax_table(phylum_physeq)[, colnames(tax_table(phylum_physeq))] <- gsub(tax_table(phylum_physeq)[, colnames(tax_table(phylum_physeq))],     pattern = "[a-z]__", replacement = "")

# check tax_table again
phyloseq::tax_table(phylum_physeq)
```

* Remove empty/unneeded samples
```{r}
# Filter phylum level phyloseq object to remove Eukaryotes
phylum_physeq <- subset_taxa(phylum_physeq, Phylum!="Eukaryota")

# prune "empty" samples
phylum_physeq <- prune_samples(sample_sums(phylum_physeq) > 0, phylum_physeq)

# Remove faecal samples
phylum_physeq <- ps_filter(phylum_physeq, type != "Faecal")
```

* Test plot (top 10)
Specifying 2:11 omits the unclassified faction from the plot, which has the highest abundance at all taxonomic levels
```{r}
TopNOTUs.Phy <- names(sort(taxa_sums(phylum_physeq), TRUE)[2:11])
phy10   <- prune_taxa(TopNOTUs.Phy, phylum_physeq)
plot_bar(phy10, fill="Phylum")
```

* Export phyloseq object for use in other scripts
```{r}
saveRDS(phylum_physeq, file = "phylum_phyloseq.rds")
```


## Class level

* Parse metaphlan data into phyloseq object
```{r}
class_physeq <- metaphlan_to_phyloseq(
  mpa_withunknown,
  metadata = metadata,
  version = 4,
  verbose = TRUE,
  tax_lvl = "Class"
)
```

* Fill in any empty cells using microviz function
```{r}
class_physeq <- class_physeq %>% tax_fix()
```

* Remove the leading characters (e.g. p__)
```{r}
# view part of the tax table
phyloseq::tax_table(class_physeq)

# replace _[A-Z] with ""
tax_table(class_physeq)[, colnames(tax_table(class_physeq))] <- gsub(tax_table(class_physeq)[, colnames(tax_table(class_physeq))],     pattern = "[a-z]__", replacement = "")

# check tax_table again
phyloseq::tax_table(class_physeq)
```

* Remove empty/unneeded samples
```{r}
# Filter class level phyloseq object to remove Eukaryotes
class_physeq <- subset_taxa(class_physeq, Class!="Eukaryota")

# prune "empty" samples
class_physeq <- prune_samples(sample_sums(class_physeq) > 0, class_physeq)

# Remove faecal samples
class_physeq <- ps_filter(class_physeq, type != "Faecal")
```

* Test plot (top 10)
```{r}
TopNOTUs.Cls <- names(sort(taxa_sums(class_physeq), TRUE)[2:11])
cls10   <- prune_taxa(TopNOTUs.Cls, class_physeq)
plot_bar(cls10, fill="Class")
```

* Export phyloseq object for use in other scripts
```{r}
saveRDS(class_physeq, file = "class_phyloseq.rds")
```

## Order level

* Parse metaphlan data into phyloseq object
```{r}
order_physeq <- metaphlan_to_phyloseq(
  mpa_withunknown,
  metadata = metadata,
  version = 4,
  verbose = TRUE,
  tax_lvl = "Order"
)
```

* Fill in any empty cells using microviz function
```{r}
order_physeq <- order_physeq %>% tax_fix()
```

* Remove the leading characters (e.g. p__)
```{r}
# view part of the tax table
phyloseq::tax_table(order_physeq)

# replace _[A-Z] with ""
tax_table(order_physeq)[, colnames(tax_table(order_physeq))] <- gsub(tax_table(order_physeq)[, colnames(tax_table(order_physeq))],     pattern = "[a-z]__", replacement = "")

# check tax_table again
phyloseq::tax_table(order_physeq)
```

* Remove empty/unneeded samples
```{r}
# Filter order level phyloseq object to remove Eukaryotes
order_physeq <- subset_taxa(order_physeq, Order!="Eukaryota")

# prune "empty" samples
order_physeq <- prune_samples(sample_sums(order_physeq) > 0, order_physeq)

# Remove faecal samples
order_physeq <- ps_filter(order_physeq, type != "Faecal")
```

* Test plot (top 10)
```{r}
TopNOTUs.Ord <- names(sort(taxa_sums(order_physeq), TRUE)[2:11])
ord10   <- prune_taxa(TopNOTUs.Ord, order_physeq)
plot_bar(ord10, fill="Order")
```

* Export phyloseq object for use in other scripts
```{r}
saveRDS(order_physeq, file = "order_phyloseq.rds")
```


## Family

* Parse metaphlan data into phyloseq object
```{r}
family_physeq <- metaphlan_to_phyloseq(
  mpa_withunknown,
  metadata = metadata,
  version = 4,
  verbose = TRUE,
  tax_lvl = "Family"
)
```

* Fill in any empty cells using microviz function
```{r}
family_physeq <- family_physeq %>% tax_fix()
```

* Remove the leading characters (e.g. p__)
```{r}
# view part of the tax table
phyloseq::tax_table(family_physeq)

# replace _[A-Z] with ""
tax_table(family_physeq)[, colnames(tax_table(family_physeq))] <- gsub(tax_table(family_physeq)[, colnames(tax_table(family_physeq))],     pattern = "[a-z]__", replacement = "")

# check tax_table again
phyloseq::tax_table(family_physeq)
```

* Remove empty/unneeded samples
```{r}
# Filter family level phyloseq object to remove Eukaryotes
family_physeq <- subset_taxa(family_physeq, Family!="Eukaryota")

# prune "empty" samples
family_physeq <- prune_samples(sample_sums(family_physeq) > 0, family_physeq)

# Remove faecal samples
family_physeq <- ps_filter(family_physeq, type != "Faecal")
```

* Test plot (top 10)
```{r}
TopNOTUs.Fam <- names(sort(taxa_sums(family_physeq), TRUE)[2:11])
fam10   <- prune_taxa(TopNOTUs.Fam, family_physeq)
plot_bar(fam10, fill="Family")
```

* Export phyloseq object for use in other scripts
```{r}
saveRDS(family_physeq, file = "family_phyloseq.rds")
```


## Genus

* Parse metaphlan data into phyloseq object
```{r}
genus_physeq <- metaphlan_to_phyloseq(
  mpa_withunknown,
  metadata = metadata,
  version = 4,
  verbose = TRUE,
  tax_lvl = "Genus"
)
```

* Fill in any empty cells using microviz function
```{r}
genus_physeq <- genus_physeq %>% tax_fix()
```

* Remove the leading characters (e.g. p__)
```{r}
# view part of the tax table
phyloseq::tax_table(genus_physeq)

# replace _[A-Z] with ""
tax_table(genus_physeq)[, colnames(tax_table(genus_physeq))] <- gsub(tax_table(genus_physeq)[, colnames(tax_table(genus_physeq))],     pattern = "[a-z]__", replacement = "")

# check tax_table again
phyloseq::tax_table(genus_physeq)
```

* Remove empty/unneeded samples
```{r}
# Filter genus level phyloseq object to remove Eukaryotes
genus_physeq <- subset_taxa(genus_physeq, Genus!="Eukaryota")

# prune "empty" samples
genus_physeq <- prune_samples(sample_sums(genus_physeq) > 0, genus_physeq)

# Remove faecal samples
genus_physeq <- ps_filter(genus_physeq, type != "Faecal")
```

* Test plot (top 10)
```{r}
TopNOTUs.Gen <- names(sort(taxa_sums(genus_physeq), TRUE)[2:11])
gen10   <- prune_taxa(TopNOTUs.Gen, genus_physeq)
plot_bar(gen10, fill="Genus")
```

* Export phyloseq object for use in other scripts
```{r}
saveRDS(genus_physeq, file = "genus_phyloseq.rds")
```

* Identify number of samples containing Segatella
```{r}
# Identify taxa that are classified as Segatella
seg_taxa <- taxa_names(subset_taxa(genus_physeq, Genus == "Segatella"))

# Subset the phyloseq object to include only Segatella
seg_physeq <- prune_taxa(seg_taxa, genus_physeq)

# Check which samples have non-zero abundance for Segatella
seg_present_samples <- sample_sums(seg_physeq) > 0

# Count the number of samples with Segatella
num_samples_with_seg <- sum(seg_present_samples)

# Output the result
cat("Number of samples with segatella:", num_samples_with_seg, "\n")
```


## Species

* Parse metaphlan data into phyloseq object
```{r}
species_physeq <- metaphlan_to_phyloseq(
  mpa_withunknown,
  metadata = metadata,
  version = 4,
  verbose = TRUE,
  tax_lvl = "Species"
)
```

* Fill in any empty cells using microviz function
```{r}
species_physeq <- species_physeq %>% tax_fix()
```

* Remove the leading characters (e.g. p__)
```{r}
# view part of the tax table
phyloseq::tax_table(species_physeq)

# replace _[A-Z] with ""
tax_table(species_physeq)[, colnames(tax_table(species_physeq))] <- gsub(tax_table(species_physeq)[, colnames(tax_table(species_physeq))],     pattern = "[a-z]__", replacement = "")

# check tax_table again
phyloseq::tax_table(species_physeq)
```

* Remove empty/unneeded samples
```{r}
# Filter species level phyloseq object to remove Eukaryotes
species_physeq <- subset_taxa(species_physeq, Species!="Eukaryota")

# prune "empty" samples
species_physeq <- prune_samples(sample_sums(species_physeq) > 0, species_physeq)

# Remove faecal samples
species_physeq <- ps_filter(species_physeq, type != "Faecal")
```

* Test plot (top 10)
```{r}
TopNOTUs.Spe <- names(sort(taxa_sums(species_physeq), TRUE)[2:11])
spe10   <- prune_taxa(TopNOTUs.Spe, species_physeq)
plot_bar(spe10, fill="Species")
```

* Export phyloseq object for use in other scripts
```{r}
saveRDS(species_physeq, file = "species_phyloseq.rds")
```