---
title: "Diversity Measures and Visualisation"
author: "Catherine Purse"
date: "2024-11-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This R markdown file contains the code needed to calculate both alpha and beta diversity measures for the metagenomic dataset explored in the manuscript "Biogeography but not age affects the captive-bred cynomolgus macaque intestinal microbiome.". It also contains the necessary code for visualisation of the diversity metrics.

* Load libraries
```{r}
library(phyloseq)
library(tidyverse)
library(microbiome)
library(grafify)
library(microViz)
library(patchwork)
library(nlme)
library(vegan)
library(pairwiseAdonis)
```

* Load Phyloseq object
```{r}
species_physeq.noEuk <- readRDS("species_phyloseq.rds")
```

* Colour palette
```{r}
cb_palette <- c("#ee8866", "#77aadd","#d3d3d3","#FDF28E","#8ED29B","#bbccee", "#AFC30A","#cc99cc","#C1E49C","#dd7788", "#009E73")
```

# Beta

* Plot ordination and perform PERMANOVA with pairwise comparisons
```{r}
## Reorder Region and Age Group factor levels
# Get sample data from the phyloseq object
sample_data_df <- as.data.frame(sample_data(species_physeq.noEuk))

# Reorder factor levels - Region
sample_data_df$type <- factor(sample_data_df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Reorder factor levels - Age Group
sample_data_df$agegroup <- factor(sample_data_df$agegroup, levels=c("Young", "Adult", "Aged"))

# Update the sample data in the phyloseq object
sample_data(species_physeq.noEuk) <- sample_data(sample_data_df)

# Ordination
set.seed(123)
mpa.ord <- ordinate(species_physeq.noEuk, "NMDS", "bray")

# Plot
mpa.ord.plot <- plot_ordination(species_physeq.noEuk, mpa.ord, type="samples", color="type", shape="agegroup") +
  theme_bw() +
  scale_shape_discrete(name = "Age Group") +
  theme(text = element_text(size = 15)) +
  geom_point(size = 4) +
  scale_color_manual(name = "Region", values = cb_palette)
print(mpa.ord.plot)

# PERMANOVA
head(sample_data(species_physeq.noEuk))
bray <- phyloseq::distance(species_physeq.noEuk, method = "bray")
sam <- data.frame(sample_data(species_physeq.noEuk))
adonis2(bray ~ agegroup + type + agegroup*type, data = sam, permutations = 999, strata = sam$id)

# Pairwise comparisons
pw.age <- pairwise.adonis2(bray ~ agegroup, data = sam, p_adjust_m = 'BH', strata = 'id')
pw.type <- pairwise.adonis2(bray ~ type, data = sam, p_adjust_m = 'BH', strata = 'agegroup')
```

* Get summary tables and significance matrix for the PERMANOVA and pairwise comp.
```{r}
# Function to extract p-values from pairwise comparisons
extract_pvalues <- function(pw_result) {
  p_values <- sapply(pw_result[-1], function(x) x$`Pr(>F)`[1])
  names(p_values) <- gsub("_vs_", "-", names(p_values))
  return(p_values)
}

# Extract p-values
age_pvalues <- extract_pvalues(pw.age)
type_pvalues <- extract_pvalues(pw.type)

# Create summary tables
age_summary <- data.frame(
  Comparison = names(age_pvalues),
  p.value = age_pvalues,
  Significant = ifelse(age_pvalues < 0.05, "*", "")
)

type_summary <- data.frame(
  Comparison = names(type_pvalues),
  p.value = type_pvalues,
  Significant = ifelse(type_pvalues < 0.05, "*", "")
)

age_summary
type_summary

# Export
write.csv(age_summary,"C:/Users/purse/OneDrive - Norwich BioScience Institutes/Results/NHP Bioinformatics/CM_Prokaryome_Paper_Oct2024/R_Analysis/Results/Ordination_Pairwise_Age_Summary.csv", row.names = TRUE)

write.csv(type_summary,"C:/Users/purse/OneDrive - Norwich BioScience Institutes/Results/NHP Bioinformatics/CM_Prokaryome_Paper_Oct2024/R_Analysis/Results/Ordination_Pairwise_Type_Summary.csv", row.names = TRUE)
```

* Export ordination plot
```{r}
folder_path <- "C:/Users/purse/OneDrive - Norwich BioScience Institutes/Results/NHP Bioinformatics/CM_Prokaryome_Paper_Oct2024/R_Analysis/Plots"
tiff(file.path(folder_path, "prokaryome_ordination.tiff"), units="in", width=6.5, height=5, res=300)
plot(mpa.ord.plot)
dev.off()
```


# Alpha
* Showing regional and age-associated differences in alpha diversity + performing statistical tests

## Regions - Chao1
```{r}
chao <- boxplot_alpha(species_physeq.noEuk,
                          index = "chao1",
                          x_var = "type",
                          )

p.chao <- chao + scale_x_discrete() +
  theme_bw() +
  labs(x="Spatial Region", y="Chao1") +
    theme(axis.text.x = element_text(size=10), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        legend.position ='none') +
  scale_fill_manual(name = "Region", values = cb_palette)
p.chao
```

# Statistical comparisons - Chao1
```{r}
chao.reg <- meta(species_physeq.noEuk)
chao.reg$diversity <- microbiome::alpha(species_physeq.noEuk, index = "all")$chao1

# Define the regions for releveling
regions <- c("D", "J", "I", "C", "PC", "DC")

# Initialize an empty list to store results
chao_results_list <- list()

# Initialize a list to store all p-values
all_p_values <- c()

# Loop over each region, relevel, fit the model, and store the results
for (region in regions) {
  chao.reg$type <- relevel(chao.reg$type, ref = region)
  
  model <- lme(diversity ~ type, random = ~ 1 | id, data = chao.reg, method = "REML")
  
  st <- summary(model)
  tabl <- as.data.frame(st$tTable)  # Convert the tTable to a data frame
  
  # Add a column to identify the reference region
  tabl$Reference <- region
  
  # Add a column to indicate significance based on p-value
  tabl$Significant <- ifelse(tabl$`p-value` < 0.05, "Yes", "No")
  
  # Store the result in the list
  chao_results_list[[region]] <- tabl
  
    # Collect all p-values (excluding the intercept)
  all_p_values <- c(all_p_values, tabl$`p-value`[-1])
}

# Apply FDR correction to all p-values
adjusted_p_values <- p.adjust(all_p_values, method = "BH")

# Initialize an index to keep track of which adjusted p-value to use
adj_p_index <- 1

# Loop through the results again to add adjusted p-values and significance
for (region in regions) {
  tabl <- chao_results_list[[region]]
  
  # Add adjusted p-values (skip the intercept)
  tabl$adjusted_p_value <- c(NA, adjusted_p_values[adj_p_index:(adj_p_index + nrow(tabl) - 2)])
  
  # Update the index
  adj_p_index <- adj_p_index + nrow(tabl) - 1
  
  # Add a column to indicate significance based on adjusted p-value
  tabl$Significant <- ifelse(tabl$adjusted_p_value < 0.05, "Yes", "No")
  
  # Update the result in the list
  chao_results_list[[region]] <- tabl
}
# Combine all the results into a single data frame
chao_results_df <- do.call(rbind, chao_results_list)

# Display the combined data frame with significance column
chao_results_df

# Create the summary table
chao_summary_table <- chao_results_df %>%
  filter(grepl("^[A-Z]+\\.type[A-Z]+$", rownames(.))) %>%  # Filter rows with type comparisons
  mutate(
    Comparison = paste0(Reference, "-", sub(".*type", "", rownames(.))),
    Significant = ifelse(Significant == "Yes", "*", "")
  ) %>%
  select(Comparison, Value, `p-value`, Significant) %>%
  arrange(Comparison)

# Create a matrix of significance
chao_sig_matrix <- chao_summary_table %>%
  mutate(From = sub("-.*", "", Comparison),
         To = sub(".*-", "", Comparison)) %>%
  select(From, To, Significant) %>%
  pivot_wider(names_from = To, values_from = Significant, values_fill = "") %>%
  column_to_rownames("From")

# Make the matrix symmetrical and remove self-comparisons
for(i in rownames(chao_sig_matrix)) {
  for(j in colnames(chao_sig_matrix)) {
    if(i == j) {
      chao_sig_matrix[i, j] = ""
    } else if(chao_sig_matrix[i, j] == "" && chao_sig_matrix[j, i] != "") {
      chao_sig_matrix[i, j] = chao_sig_matrix[j, i]
    }
  }
}

# Reorder matrix
desired_order <- c("D", "J", "I", "C", "PC", "DC")
chao_sig_matrix <- chao_sig_matrix %>%
  select(all_of(desired_order)) %>%
  slice(match(desired_order, rownames(.)))

# Print the results
print(chao_summary_table)
print(chao_sig_matrix)
```


# Regions - Inverse Simpson
```{r}
# Diversity Inverse Simpson
inverse_simpson <- boxplot_alpha(species_physeq.noEuk,
                          index = "inverse_simpson",
                          x_var = "type",
                          )

p.inverse_simpson <- inverse_simpson + 
  scale_x_discrete() +
  theme_bw() +
  labs(x="Spatial Region", y="Inverse Simpson") +
  theme(axis.text.x = element_text(size=10), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        legend.position ='none') +
  scale_fill_manual(name = "Region", values = cb_palette)
p.inverse_simpson
```

# Statistical comparisons - Diversity Inverse Simpson
```{r}
inverse_simpson.reg <- meta(species_physeq.noEuk)
inverse_simpson.reg$diversity <- microbiome::alpha(species_physeq.noEuk, index = "all")$diversity_inverse_simpson

# Define the regions for releveling
regions <- c("D", "J", "I", "C", "PC", "DC")

# Initialize an empty list to store results
inverse_simpson_results_list <- list()

# Initialize a list to store all p-values
all_p_values <- c()

# Loop over each region, relevel, fit the model, and store the results
for (region in regions) {
  inverse_simpson.reg$type <- relevel(inverse_simpson.reg$type, ref = region)
  
  model <- lme(diversity ~ type, random = ~ 1 | id, data = inverse_simpson.reg, method = "REML")
  
  st <- summary(model)
  tabl <- as.data.frame(st$tTable)  # Convert the tTable to a data frame
  
  # Add a column to identify the reference region
  tabl$Reference <- region
  
  # Add a column to indicate significance based on p-value
  tabl$Significant <- ifelse(tabl$`p-value` < 0.05, "Yes", "No")
  
  # Store the result in the list
  inverse_simpson_results_list[[region]] <- tabl
  
  # Collect all p-values (excluding the intercept)
  all_p_values <- c(all_p_values, tabl$`p-value`[-1])
}

# Apply FDR correction to all p-values
adjusted_p_values <- p.adjust(all_p_values, method = "BH")

# Initialize an index to keep track of which adjusted p-value to use
adj_p_index <- 1

# Loop through the results again to add adjusted p-values and significance
for (region in regions) {
  tabl <- inverse_simpson_results_list[[region]]
  
  # Add adjusted p-values (skip the intercept)
  tabl$adjusted_p_value <- c(NA, adjusted_p_values[adj_p_index:(adj_p_index + nrow(tabl) - 2)])
  
  # Update the index
  adj_p_index <- adj_p_index + nrow(tabl) - 1
  
  # Add a column to indicate significance based on adjusted p-value
  tabl$Significant <- ifelse(tabl$adjusted_p_value < 0.05, "Yes", "No")
  
  # Update the result in the list
  inverse_simpson_results_list[[region]] <- tabl
}

# Combine all the results into a single data frame
inverse_simpson_results_df <- do.call(rbind, inverse_simpson_results_list)

# Display the combined data frame with significance column
inverse_simpson_results_df

# Create the summary table
inverse_simpson_summary_table <- inverse_simpson_results_df %>%
  filter(grepl("^[A-Z]+\\.type[A-Z]+$", rownames(.))) %>%  # Filter rows with type comparisons
  mutate(
    Comparison = paste0(Reference, "-", sub(".*type", "", rownames(.))),
    Significant = ifelse(Significant == "Yes", "*", "")
  ) %>%
  select(Comparison, Value, `p-value`, Significant) %>%
  arrange(Comparison)

# Create a matrix of significance
inverse_simpson_sig_matrix <- inverse_simpson_summary_table %>%
  mutate(From = sub("-.*", "", Comparison),
         To = sub(".*-", "", Comparison)) %>%
  select(From, To, Significant) %>%
  pivot_wider(names_from = To, values_from = Significant, values_fill = "") %>%
  column_to_rownames("From")

# Create the significance matrix
inverse_simpson_sig_matrix <- inverse_simpson_summary_table %>%
  mutate(From = sub("-.*", "", Comparison),
         To = sub(".*-", "", Comparison)) %>%
  select(From, To, Significant) %>%
  pivot_wider(names_from = To, values_from = Significant, values_fill = "") %>%
  column_to_rownames("From")

# Reorder matrix
desired_order <- c("D", "J", "I", "C", "PC", "DC")
inverse_simpson_sig_matrix <- inverse_simpson_sig_matrix[desired_order, desired_order]

# Fill diagonal with empty strings
diag(inverse_simpson_sig_matrix) <- ""

# Print the results
print(inverse_simpson_summary_table)
print(inverse_simpson_sig_matrix)
```


# Regions - Shannon
```{r}
# Regions - Shannon
shannon <- boxplot_alpha(species_physeq.noEuk,
                          index = "shannon",
                          x_var = "type",
                          )

p.shannon <- shannon + scale_x_discrete() +
  theme_bw() +
  labs(x="Spatial Region", y="Shannon") +
    theme(axis.text.x = element_text(size=10), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        legend.position ='none') +
  scale_fill_manual(name = "Region", values = cb_palette)
p.shannon
```

# Statistical comparisons - Shannon
```{r}
sh.reg <- meta(species_physeq.noEuk)
sh.reg$diversity <- microbiome::alpha(species_physeq.noEuk, index = "all")$diversity_shannon

# Define the regions for releveling
regions <- c("D", "J", "I", "C", "PC", "DC")

# Initialize an empty list to store results
diversity_shannon_results_list <- list()

# Initialize a list to store all p-values
all_p_values <- c()

# Loop over each region, relevel, fit the model, and store the results
for (region in regions) {
  sh.reg$type <- relevel(sh.reg$type, ref = region)
  
  model <- lme(diversity ~ type, random = ~ 1 | id, data = sh.reg, method = "REML")
  
  st <- summary(model)
  tabl <- as.data.frame(st$tTable)  # Convert the tTable to a data frame
  
  # Add a column to identify the reference region
  tabl$Reference <- region
  
  # Add a column to indicate significance based on p-value
  tabl$Significant <- ifelse(tabl$`p-value` < 0.05, "Yes", "No")
  
  # Store the result in the list
  diversity_shannon_results_list[[region]] <- tabl
  
  # Collect all p-values (excluding the intercept)
  all_p_values <- c(all_p_values, tabl$`p-value`[-1])
}

# Apply FDR correction to all p-values
adjusted_p_values <- p.adjust(all_p_values, method = "BH")

# Initialize an index to keep track of which adjusted p-value to use
adj_p_index <- 1

# Loop through the results again to add adjusted p-values and significance
for (region in regions) {
  tabl <- diversity_shannon_results_list[[region]]
  
  # Add adjusted p-values (skip the intercept)
  tabl$adjusted_p_value <- c(NA, adjusted_p_values[adj_p_index:(adj_p_index + nrow(tabl) - 2)])
  
  # Update the index
  adj_p_index <- adj_p_index + nrow(tabl) - 1
  
  # Add a column to indicate significance based on adjusted p-value
  tabl$Significant <- ifelse(tabl$adjusted_p_value < 0.05, "Yes", "No")
  
  # Update the result in the list
  diversity_shannon_results_list[[region]] <- tabl
}
# Combine all the results into a single data frame
shannon_results_df <- do.call(rbind, diversity_shannon_results_list)

# Display the combined data frame with significance column
shannon_results_df

# Create the summary table
shannon_summary_table <- shannon_results_df %>%
  filter(grepl("^[A-Z]+\\.type[A-Z]+$", rownames(.))) %>%  # Filter rows with type comparisons
  mutate(
    Comparison = paste0(Reference, "-", sub(".*type", "", rownames(.))),
    Significant = ifelse(Significant == "Yes", "*", "")
  ) %>%
  select(Comparison, Value, `p-value`, Significant) %>%
  arrange(Comparison)

# Create a matrix of significance
shannon_sig_matrix <- shannon_summary_table %>%
  mutate(From = sub("-.*", "", Comparison),
         To = sub(".*-", "", Comparison)) %>%
  select(From, To, Significant) %>%
  pivot_wider(names_from = To, values_from = Significant, values_fill = "") %>%
  column_to_rownames("From")

# Make the matrix symmetrical and remove self-comparisons
for(i in rownames(shannon_sig_matrix)) {
  for(j in colnames(shannon_sig_matrix)) {
    if(i == j) {
      shannon_sig_matrix[i, j] = ""
    } else if(shannon_sig_matrix[i, j] == "" && shannon_sig_matrix[j, i] != "") {
      shannon_sig_matrix[i, j] = shannon_sig_matrix[j, i]
    }
  }
}

# Reorder matrix
desired_order <- c("D", "J", "I", "C", "PC", "DC")
shannon_sig_matrix <- shannon_sig_matrix %>%
  select(all_of(desired_order)) %>%
  slice(match(desired_order, rownames(.)))

# Print the results
print(shannon_summary_table)
print(shannon_sig_matrix)
```

# Regions - Patchwork
* Export
```{r}
reg.alpha_patchwork <- p.chao / p.inverse_simpson / p.shannon +
   plot_layout(ncol = 2, guides = "collect", axis_titles = "collect")
reg.alpha_patchwork

tiff(file.path(folder_path, "prokaryome_alpha_regions.tiff"), units="in", width=8, height=6, res=300)
plot(reg.alpha_patchwork)
dev.off()
```


# Age - Chao1
```{r}
chao <- meta(species_physeq.noEuk)
chao$diversity <- microbiome::alpha(species_physeq.noEuk, index = "all")$chao1

# Chao1 - Linear regression for all regions
# Plot
c_age_chao <- ggplot(chao,
       aes(x=age,
           y=diversity))+
  geom_point()+
  geom_smooth(method = "lm",
              se = FALSE,
              colour = "#77aadd")+
  theme_light() +
  theme(axis.text.x = element_text(angle=0, size=10), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold")) +
  facet_grid("type") +
  xlab("Age (Years)") + 
  ylab("Chao1")

# Linear regressions per sample site

# Initialize an empty list to store results
a_diversity_chao_results_list <- list()

# Initialize a list to store all p-values
all_p_values <- c()

# Define the regions for releveling
regions <- c("D", "J", "I", "C", "PC", "DC")

# Loop over each region, relevel, fit the model, and store the results
for (region in regions) {
  # Subset the data for the current region
  chao_temp <- subset(chao, type == region)
  
  model <- lm(diversity ~ age,
                   data = chao_temp)
  
  model_summary <- summary(model)
  coef_matrix <- model_summary$coefficients
  coef_df <- as.data.frame(coef_matrix)

  
  # Add a column to identify the reference region
  coef_df$Reference <- region
  
  # Add a column to indicate significance based on p-value
  coef_df$Significant <- ifelse(coef_df$`Pr(>|t|)` < 0.05, "Yes", "No")
  
  # Store the result in the list
  a_diversity_chao_results_list[[region]] <- coef_df
  
  # Collect all p-values (excluding the intercept)
  all_p_values <- c(all_p_values, coef_df$`Pr(>|t|)`[-1])
}


# Combine all the results into a single data frame
chao_results_df <- do.call(rbind, a_diversity_chao_results_list)

# Display the combined data frame with significance column
chao_results_df

# Create the summary table
chao_summary_table <- chao_results_df %>%
  filter(grepl("^[A-Z]+\\.age", rownames(.))) %>%  # Filter rows with type comparisons
  mutate(
    Comparison = paste0(Reference, "-age"),
    Significant = ifelse(Significant == "Yes", "*", "")
  ) %>%
  select(Comparison, Estimate, `Pr(>|t|)`, Significant) %>%
  arrange(Comparison)

# Create a matrix of significance
chao_sig_matrix <- chao_summary_table %>%
  mutate(From = sub("-.*", "", Comparison),
         To = sub(".*-", "", Comparison)) %>%
  select(From, To, Significant) %>%
  pivot_wider(names_from = To, values_from = Significant, values_fill = "") %>%
  column_to_rownames("From")

# Reorder matrix
desired_order <- c("D", "J", "I", "C", "PC", "DC")
chao_sig_matrix[desired_order,, drop = FALSE]

# Print the results
print(chao_summary_table)
print(chao_sig_matrix)

tiff(file.path(folder_path, "chao1_con_age.tiff"), units="in", width=4, height=4, res=300)
plot(c_age_chao)
dev.off()
```  

# Age - Shannon
```{r}
shannon <- meta(species_physeq.noEuk)
shannon$diversity <- microbiome::alpha(species_physeq.noEuk, index = "all")$diversity_shannon 

# Shannon - Linear regression for all regions
# Plot
c_age_shannon <- ggplot(shannon,
       aes(x = age,
           y = diversity)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE,
              colour = "#77aadd") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 0, size = 10), 
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background = element_rect(fill = "grey"),
        strip.text = element_text(size = 12, face = "bold")) +
  facet_grid("type") +
  xlab("Age (Years)") + 
  ylab("Shannon")

# Linear regressions per sample site

# Initialize an empty list to store results
a_diversity_shannon_results_list <- list()

# Initialize a list to store all p-values
all_p_values <- c()

# Define the regions for releveling
regions <- c("D", "J", "I", "C", "PC", "DC")

# Loop over each region, relevel, fit the model, and store the results
for (region in regions) {
  # Subset the data for the current region
  shannon_temp <- subset(shannon, type == region)
  
  model <- lm(diversity ~ age,
                   data = shannon_temp)
  
  model_summary <- summary(model)
  coef_matrix <- model_summary$coefficients
  coef_df <- as.data.frame(coef_matrix)

  
  # Add a column to identify the reference region
  coef_df$Reference <- region
  
  # Add a column to indicate significance based on p-value
  coef_df$Significant <- ifelse(coef_df$`Pr(>|t|)` < 0.05, "Yes", "No")
  
  # Store the result in the list
  a_diversity_shannon_results_list[[region]] <- coef_df
  
  # Collect all p-values (excluding the intercept)
  all_p_values <- c(all_p_values, coef_df$`Pr(>|t|)`[-1])
}


# Combine all the results into a single data frame
shannon_results_df <- do.call(rbind, a_diversity_shannon_results_list)

# Display the combined data frame with significance column
shannon_results_df

# Create the summary table
shannon_summary_table <- shannon_results_df %>%
  filter(grepl("^[A-Z]+\\.age", rownames(.))) %>%  # Filter rows with type comparisons
  mutate(
    Comparison = paste0(Reference, "-age"),
    Significant = ifelse(Significant == "Yes", "*", "")
  ) %>%
  select(Comparison, Estimate, `Pr(>|t|)`, Significant) %>%
  arrange(Comparison)

# Create a matrix of significance
shannon_sig_matrix <- shannon_summary_table %>%
  mutate(From = sub("-.*", "", Comparison),
         To = sub(".*-", "", Comparison)) %>%
  select(From, To, Significant) %>%
  pivot_wider(names_from = To, values_from = Significant, values_fill = "") %>%
  column_to_rownames("From")

# Reorder matrix
desired_order <- c("D", "J", "I", "C", "PC", "DC")
shannon_sig_matrix <- shannon_sig_matrix[desired_order,, drop = FALSE]

# Print the results
print(shannon_summary_table)
print(shannon_sig_matrix)

tiff(file.path(folder_path, "shannon_con_age.tiff"), units="in", width=4, height=4, res=300)
plot(c_age_shannon)
dev.off()
```


# Age - Inverse Simpson
```{r}
inverse_simpson <- meta(species_physeq.noEuk)
inverse_simpson$diversity <- microbiome::alpha(species_physeq.noEuk, index = "all")$diversity_inverse_simpson

# Inverse Simpson - Linear regression for all regions
# Plot
c_age_invsimpson <- ggplot(inverse_simpson,
       aes(x = age,
           y = diversity)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE,
              colour = "#77aadd") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 0, size = 10), 
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background = element_rect(fill = "grey"),
        strip.text = element_text(size = 12, face = "bold")) +
  facet_grid("type") +
  xlab("Age (Years)") + 
  ylab("Inverse Simpson")

# Linear regressions per sample site

# Initialize an empty list to store results
a_diversity_inverse_simpson_results_list <- list()

# Initialize a list to store all p-values
all_p_values <- c()

# Define the regions for releveling
regions <- c("D", "J", "I", "C", "PC", "DC")

# Loop over each region, relevel, fit the model, and store the results
for (region in regions) {
  # Subset the data for the current region
  inverse_simpson_temp <- subset(inverse_simpson, type == region)
  
  model <- lm(diversity ~ age,
              data = inverse_simpson_temp)
  
  model_summary <- summary(model)
  coef_matrix <- model_summary$coefficients
  coef_df <- as.data.frame(coef_matrix)

  
  # Add a column to identify the reference region
  coef_df$Reference <- region
  
  # Add a column to indicate significance based on p-value
  coef_df$Significant <- ifelse(coef_df$`Pr(>|t|)` < 0.05, "Yes", "No")
  
  # Store the result in the list
  a_diversity_inverse_simpson_results_list[[region]] <- coef_df
  
  # Collect all p-values (excluding the intercept)
  all_p_values <- c(all_p_values, coef_df$`Pr(>|t|)`[-1])
}


# Combine all the results into a single data frame
inverse_simpson_results_df <- do.call(rbind, a_diversity_inverse_simpson_results_list)

# Display the combined data frame with significance column
inverse_simpson_results_df

# Create the summary table
inverse_simpson_summary_table <- inverse_simpson_results_df %>%
  filter(grepl("^[A-Z]+\\.age", rownames(.))) %>%  # Filter rows with type comparisons
  mutate(
    Comparison = paste0(Reference, "-age"),
    Significant = ifelse(Significant == "Yes", "*", "")
  ) %>%
  select(Comparison, Estimate, `Pr(>|t|)`, Significant) %>%
  arrange(Comparison)

# Create a matrix of significance
inverse_simpson_sig_matrix <- inverse_simpson_summary_table %>%
  mutate(From = sub("-.*", "", Comparison),
         To = sub(".*-", "", Comparison)) %>%
  select(From, To, Significant) %>%
  pivot_wider(names_from = To, values_from = Significant, values_fill = "") %>%
  column_to_rownames("From")

# Reorder matrix
desired_order <- c("D", "J", "I", "C", "PC", "DC")
inverse_simpson_sig_matrix <- inverse_simpson_sig_matrix[desired_order,, drop = FALSE]

# Print the results
print(inverse_simpson_summary_table)
print(inverse_simpson_sig_matrix)


tiff(file.path(folder_path, "invSimpson_con_age.tiff"), units="in", width=4, height=4, res=300)
plot(c_age_invsimpson)
dev.off()
```

Note: any lines of code that contained file paths (i.e. for export of .csv files) have been removed in the interests of security.