---
title: "Novel SGB Visualisation"
author: "Catherine Purse"
date: "2024-11-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* Load libraries
```{r}
library(phyloseq)
library(tidyverse)
library(microbiome)
library(grafify)
library(microViz)
library(patchwork)
```

* Load otu_table, metadata, and taxa
```{r}
bins <- read.csv("novelSGBs_mergedabundance.csv")
tax <- read.csv("novelSGBs_taxtable.csv")
meta <- read.csv("SGBs_metadata_6189_Oct24.csv")
```

## Create phyloseq object

* Alter bins/tax/meta dataframes so they are in the right format
```{r}
#create bin table in the right format (turn bin column into row names)
bins <- bins %>% remove_rownames %>% column_to_rownames(var = "Genomic_bins")

#do the same for taxonomy
tax <- tax %>% remove_rownames %>% column_to_rownames(var = "user_genome")

#and for metadata (uses a different method but does the same thing)
meta <- meta %>%
  tibble::column_to_rownames("sample")

#strip white space from all (to make sure names match across sheets)
tax <- tax %>%
  mutate(across(where(is.character), str_trim))
bins <- bins %>%
  mutate(across(where(is.character), str_trim))
meta <- meta %>%
  mutate(across(where(is.character), str_trim))
```

* Convert otu_table (bins) and tax_table (tax) into matrices (required prior to constructing phyloseq object)
```{r}
bins_mat <- as.matrix(bins)
tax_mat <- as.matrix(tax)
```

* Construct phyloseq object
```{r}
bins_ps <- otu_table(bins_mat, taxa_are_rows = TRUE)
tax_ps <- tax_table(tax_mat)
meta_ps <- sample_data(meta)

ps <- phyloseq(bins_ps, tax_ps, meta_ps)
```

* Remove any SGBs identifiable to species level from the data
```{r}
# Remove rows where 'species' is not empty
ps <- subset_taxa(ps, is.na(Species) | Species == "")

# Check data
unique(tax_table(ps)[,"Species"] )
```

## Clean taxonomy table

* Remove prefixes from tax_table
```{r}
#view part of the tax table
phyloseq::tax_table(ps)[1:10, 1:2]

#replace _[A-Z] with ""
tax_table(ps)[, colnames(tax_table(ps))] <- gsub(tax_table(ps)[, colnames(tax_table(ps))],     pattern = "[a-z]__", replacement = "")

#check tax_table again
phyloseq::tax_table(ps)[1:10, 1:2]
```
You can see that, for example, "Firmicutes_A" is replaced with just "Firmicutes".

We can check that there are no other unusual naming conventions by looking at the unique taxa in each rank:
```{r}
unique(tax_table(ps)[,"Genus"] )
```

Next we will fill in any missing taxonomic information, using a function from the package microViz. This is especially apparent in the Species rank, where many entries appear to be missing. This would cause problems if we try to aggregate at different levels, so we will use tax_fix to fill in these values with those from the next higher taxonomic rank.

```{r}
ps_fix <- ps %>% tax_fix()
```

* Export phyloseq object for use in other scripts
```{r}
saveRDS(ps_fix, file = "sgb_phyloseq.rds")
```

* Remove empty samples
```{r}
any(sample_sums(ps_fix) == 0) #Result is TRUE
# prune "empty" samples
ps_fix <- prune_samples(sample_sums(ps_fix) > 0, ps_fix)
```


* Create a table of number of features for each Phylum
```{r}
rank_names(ps_fix)
unique(tax_table(ps_fix)[,"Phylum"] )
table(tax_table(ps_fix)[, "Phylum"], exclude = NULL)
```

* Explore feature prevalence, defined here as the number of samples in which a taxon appears at least once.
```{r}
# Compute prevalence of each feature, store as data.frame
prevdf = apply(X = otu_table(ps_fix),
               MARGIN = ifelse(taxa_are_rows(ps_fix), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})
# Add taxonomy and total read counts to this data.frame
prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps_fix),
                    tax_table(ps_fix))

#create empty temporary directory
temp_dir <- tempdir()

# Use ddply with the specified temporary directory
result <- plyr::ddply(prevdf, "Phylum", function(df1) {
  data.frame(MeanPrevalence = mean(df1$Prevalence),
             SumPrevalence = sum(df1$Prevalence))
}, .parallel = FALSE, .paropts = list(tmpdir = temp_dir))

# Now, merge the two data frames by the "Phylum" column
merged_data <- merge(prevdf, result, by = "Phylum")

# Subset to the remaining phyla
prevdf1 = subset(merged_data, Phylum %in% get_taxa_unique(ps_fix, "Phylum"))
phyla_prev <- ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(ps_fix),color=Phylum)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme_bw() +
  theme(legend.position="none") +
  theme_bw()
  
phyla_prev
```

# SGB prevalence - Kingdom/Phylum level
```{r}
# Count the number of OTUs for each phylum
otu_counts <- aggregate(Prevalence ~ Phylum, data = prevdf1, FUN = length)

# Rename the columns for clarity
names(otu_counts)[2] <- "NumberOfSGBs"

# Merge the counts back into the original dataframe
data <- merge(prevdf1, otu_counts, by = "Phylum")

agg_data <- data %>%
  group_by(Phylum, Kingdom) %>%
  dplyr::summarise(NumberOfSGBs = n(), .groups = 'drop')

# Reorder phylum factor based on NumberOfOTUs
agg_data <- agg_data %>%
  mutate(Phylum = factor(Phylum, levels = agg_data %>%
                           arrange(desc(NumberOfSGBs)) %>%
                           pull(Phylum)))

# Plot using the aggregated data
sgb.prev <- ggplot(agg_data, aes(x = Phylum, y = NumberOfSGBs, fill = Kingdom)) +
  geom_bar(stat = "identity") +
  labs(x = "Phylum", y = "Number of SGBs") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust=0.95)) +
  geom_text(aes(label = NumberOfSGBs), vjust = -0.5, color = "black") +
  scale_fill_manual(values = c("Bacteria" = "#77aadd", "Archaea" = "#ee8866")) + 
  ylim(0,120)

print(sgb.prev)

tiff(file.path(folder_path, "sgb__kingdom_phylum.tiff"), units="in", width=5, height=2.5, res=300)
plot(sgb.prev)
dev.off()
```


# SGB prevalence - Class/Kingdom level
```{r}
# Count the number of OTUs for each phylum
class_counts <- aggregate(Prevalence ~ Class, data = prevdf1, FUN = length)

# Rename the columns for clarity
names(class_counts)[2] <- "NumberOfSGBs"

# Merge the counts back into the original dataframe
class.data <- merge(prevdf1, class_counts, by = "Class")

class.agg_data <- class.data %>%
  group_by(Kingdom, Class) %>%
  dplyr::summarise(NumberOfSGBs = n(), .groups = 'drop')

# Reorder phylum factor based on NumberOfOTUs
class.agg_data <- class.agg_data %>%
  mutate(Class = factor(Class, levels = class.agg_data %>%
                           arrange(desc(NumberOfSGBs)) %>%
                           pull(Class)))


cl.prev <- ggplot(class.agg_data, aes(x = Class, y = NumberOfSGBs, fill = Kingdom)) +
  geom_bar(stat = "identity") +
  labs(x = "Class", y = "Number of SGBs") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95)
        ) +
  geom_text(aes(label = NumberOfSGBs), vjust = -0.5, color = "black") +
  scale_fill_manual(values = c("Bacteria" = "#77aadd", "Archaea" = "#ee8866")) +
  ylim(0,120)
cl.prev

```

# SGB prevalence - Order/Kingdom level
```{r}
# Count the number of OTUs for each phylum
order_counts <- aggregate(Prevalence ~ Order, data = prevdf1, FUN = length)

# Rename the columns for clarity
names(order_counts)[2] <- "NumberOfSGBs"

# Merge the counts back into the original dataframe
ord.data <- merge(prevdf1, order_counts, by = "Order")

ord.agg_data <- ord.data %>%
  group_by(Order, Kingdom) %>%
  dplyr::summarise(NumberOfSGBs = n(), .groups = 'drop')


# Reorder phylum factor based on NumberOfOTUs
ord.agg_data <- ord.agg_data %>%
  mutate(Order = factor(Order, levels = ord.agg_data %>%
                           arrange(desc(NumberOfSGBs)) %>%
                           pull(Order)))

# Plot using the aggregated data
co.prev <- ggplot(ord.agg_data, aes(x = Order, y = NumberOfSGBs, fill = Kingdom)) +
  geom_bar(stat = "identity") +
  labs(x = "Order", y = "Number of SGBs") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95)
        ) +
  geom_text(aes(label = NumberOfSGBs), vjust = -0.5, color = "black") +
  scale_fill_manual(values = c("Bacteria" = "#77aadd", "Archaea" = "#ee8866")) +
  ylim(0,70)
co.prev

print(co.prev)
```

# SGB prevalence - Family/Kingdom level
```{r}
# Count the number of OTUs for each phylum
fam_counts <- aggregate(Prevalence ~ Family, data = prevdf1, FUN = length)

# Rename the columns for clarity
names(fam_counts)[2] <- "NumberOfSGBs"

# Merge the counts back into the original dataframe
fam.data <- merge(prevdf1, fam_counts, by = "Family")

fam.agg_data <- fam.data %>%
  group_by(Family, Kingdom) %>%
  dplyr::summarise(NumberOfSGBs = n(), .groups = 'drop')

# Reorder phylum factor based on NumberOfOTUs
fam.agg_data <- fam.agg_data %>%
  mutate(Family = factor(Family, levels = fam.agg_data %>%
                           arrange(desc(NumberOfSGBs)) %>%
                           pull(Family)))

# Plot using the aggregated data
gf.prev <- ggplot(fam.agg_data, aes(x = Family, y = NumberOfSGBs, fill = Kingdom)) +
  geom_bar(stat = "identity") +
  labs(x = "Family", y = "Number of SGBs") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust=0.95)) +
  geom_text(aes(label = NumberOfSGBs), vjust = -0.5, color = "black") +
  scale_fill_grafify(palette = "kelly", reverse = TRUE) +
  scale_fill_manual(values = c("Bacteria" = "#77aadd", "Archaea" = "#ee8866")) +
  ylim(0,70)


print(gf.prev)

sgb_pw <- sgb.prev / cl.prev / co.prev / gf.prev +
   plot_layout(ncol = 1, guides = "collect", axis_titles = "collect") +
  plot_annotation(tag_levels = 'A') +
  theme(plot.tag.position = c(0, 1),
      plot.tag = element_text(size = 12, hjust = 0, vjust = 0))
sgb_pw

folder_path <- "C:/Users/purse/OneDrive - Norwich BioScience Institutes/Results/NHP Bioinformatics/CM_Prokaryome_Paper_Oct2024/R_Analysis/Plots"
tiff(file.path(folder_path, "sgb_prev_pw.tiff"), units="in", width=10, height=12, res=300)
plot(sgb_pw)
dev.off()
```

# Genus Heatmap - colour by family, segment by type
```{r}
#Melt ps.b to df
ps.df <- psmelt(ps_fix)

# make region a factor in the df (i.e. change order of facets in next step)
ps.df$type <- factor(ps.df$type, levels=c("D", "J", "I", "C", "PC", "DC", "F"))

# Order labels by age (using a unique identifier to account for duplicates in df)
ps.df <- ps.df[order(ps.df$age, ps.df$age_label), ]
ps.df$age_label <- factor(ps.df$age_label, levels = unique(ps.df$age_label))

labels <- c(D = "D", J = "Jejunum", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon", F = "Faecal")

sgb_gen <- ggplot(ps.df, aes(age_label, Genus, fill= Abundance, alpha = Abundance)) + 
  theme_light() +
  geom_tile(colour = "gray50") +
     scale_fill_gradient(
    low = "#CAEBC3", 
  high = "#008856") +
  scale_alpha_identity(guide = "none") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(angle = 90, size=10, hjust = 0.95, vjust = 0.4),
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face="bold", colour="white"),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        ) +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  labs(x = "ID", y = "Genus", fill = "Abundance (CPM)")
sgb_gen

tiff(file.path(folder_path, "sgb_gen.tiff"), units="in", width=14.5, height=14, res=300)
plot(sgb_gen)
dev.off()
```
