---
title: "Taxa Abundance Figures"
author: "Catherine Purse"
date: "2024-11-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This R Markdown file contains the code needed to create figures for visualising the abundance of taxa for the manuscript 'Biogeography but not age affects the captive-bred cynomolgus macaque intestinal microbiome.'. It has the code to explore the top 10 most abundant taxa for both the bacterial and archaeal fractions of the intestinal microbiome. 


* Load libraries
```{r}
library(phyloseq)
library(tidyverse)
library(microbiome)
library(microViz)
library(patchwork)
library(MicEco)
```

* Colour palette
```{r}
cb_palette <- c("#ee8866", "#77aadd","#d3d3d3","#FDF28E","#8ED29B","#bbccee", "#AFC30A","#cc99cc","#C1E49C","#dd7788", "#009E73")
```

# Kingdom

* Load RDS file
```{r}
king <- readRDS("kingdom_phyloseq.rds")
```

## Faceted bar chart - Kingdom
```{r}
# melt into df
king.df <- psmelt(king)

# remove underscores between words in Phylum column
king.df$Kingdom <- gsub("_", " ", king.df$Kingdom)

# make region a factor in the df (i.e. change order of facets in next step)
king.df$type <- factor(king.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "J", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
king.df <- king.df[order(king.df$age, king.df$age_label), ]
king.df$age_label <- factor(king.df$age_label, levels = unique(king.df$age_label))


# plot
king_plot <- ggplot(king.df, aes(x = age_label, y = Abundance, fill = Kingdom)) +
  theme_light() +
  geom_col() +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=7, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=9),
        legend.title = element_text(size=12),
        legend.position = "right") +
  scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
king_plot
```


# Phyla

* Read RDS file
```{r}
phylum <- readRDS("phylum_phyloseq.rds")
```

* Filter for top 10 (excluding 'unclassified')
```{r}
TopNOTUs.Phy <- names(sort(taxa_sums(phylum), TRUE)[2:11])
phy10   <- prune_taxa(TopNOTUs.Phy, phylum)
```

## Faceted bar chart - Phyla
```{r}
# melt into df
phy10.df <- psmelt(phy10)

# remove underscores between words in Phylum column
phy10.df$Phylum <- gsub("_", " ", phy10.df$Phylum)

# make region a factor in the df (i.e. change order of facets in next step)
phy10.df$type <- factor(phy10.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "Jejunum", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
phy10.df <- phy10.df[order(phy10.df$age, phy10.df$age_label), ]
phy10.df$age_label <- factor(phy10.df$age_label, levels = unique(phy10.df$age_label))

# plot
phy_plot <- ggplot(phy10.df, aes(x = age_label, y = Abundance, fill = Phylum)) +
  theme_light() +
  geom_col() +
  ggtitle("Phylum") +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=9, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=10),
        legend.title = element_text(size=12),
        legend.position = "bottom") +
  scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
phy_plot
```


# Class

* Read RDS file
```{r}
class <- readRDS("class_phyloseq.rds")
```

* Filter for top 10 (excluding 'unclassified')
```{r}
TopNOTUs.Class <- names(sort(taxa_sums(class), TRUE)[2:11])
class10   <- prune_taxa(TopNOTUs.Class, class)
```

* Faceted bar chart - Class
```{r}
# melt into df
class10.df <- psmelt(class10)

# remove underscores between words in Class column
class10.df$Class <- gsub("_", " ", class10.df$Class)

# make region a factor in the df (i.e. change order of facets in next step)
class10.df$type <- factor(class10.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "Jejunum", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
class10.df <- class10.df[order(class10.df$age, class10.df$age_label), ]
class10.df$age_label <- factor(class10.df$age_label, levels = unique(class10.df$age_label))

# plot
class_plot <- ggplot(class10.df, aes(x = id, y = Abundance, fill = Class)) +
  theme_light() +
  geom_col() +
  ggtitle("Class") +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=10, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=10),
        legend.title = element_text(size=12),
        legend.position = "bottom") +
   scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
class_plot
```

# Order

* Read RDS file
```{r}
order <- readRDS("order_phyloseq.rds")
```

* Filter for top 10 (excluding 'unclassified')
```{r}
TopNOTUs.Order <- names(sort(taxa_sums(order), TRUE)[2:11])
order10   <- prune_taxa(TopNOTUs.Order, order)
```

## Order Rel. Abundance

```{r}
# melt into df
order10.df <- psmelt(order10)

# remove underscores between words in Order column
order10.df$Order <- gsub("_", " ", order10.df$Order)

# make region a factor in the df (i.e. change order of facets in next step)
order10.df$type <- factor(order10.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "Jejunum", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
order10.df <- order10.df[order(order10.df$age, order10.df$age_label), ]
order10.df$age_label <- factor(order10.df$age_label, levels = unique(order10.df$age_label))

# plot
order_plot <- ggplot(order10.df, aes(x = id, y = Abundance, fill = Order)) +
  theme_light() +
  geom_col() +
  ggtitle("Order") +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=10, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=10),
        legend.title = element_text(size=12),
        legend.position = "bottom") +
  scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
order_plot
```


# Family

* Read RDS file
```{r}
family <- readRDS("family_phyloseq.rds")
```

* Filter for top 10 taxa (excluding 'unclassified')
```{r}
TopNOTUs.Family <- names(sort(taxa_sums(family), TRUE)[2:11])
family10   <- prune_taxa(TopNOTUs.Family, family)
```

## Family Rel. Abundance
```{r}
# melt into df
family10.df <- psmelt(family10)

# remove underscores between words in Family column
family10.df$Family <- gsub("_", " ", family10.df$Family)

# make region a factor in the df (i.e. change order of facets in next step)
family10.df$type <- factor(family10.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "Jejunum", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
family10.df <- family10.df[order(family10.df$age, family10.df$age_label), ]
family10.df$age_label <- factor(family10.df$age_label, levels = unique(family10.df$age_label))


# plot
family_plot <- ggplot(family10.df, aes(x = id, y = Abundance, fill = Family)) +
  theme_light() +
  geom_col() +
  ggtitle("Family") +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=10, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=9),
        legend.title = element_text(size=11),
        legend.position = "bottom") +
  scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
family_plot
```

# Genus

* Read RDS file
```{r}
genus <- readRDS("genus_phyloseq.rds")
```

* Filter for top 10 taxa (excluding 'unclassified')
```{r}
TopNOTUs.Genus <- names(sort(taxa_sums(genus), TRUE)[2:11])
genus10   <- prune_taxa(TopNOTUs.Genus, genus)
```

## Genus Rel. Abundance
```{r}
# melt into df
genus10.df <- psmelt(genus10)

# remove underscores between words in Genus column
genus10.df$Genus <- gsub("_", " ", genus10.df$Genus)

# make region a factor in the df (i.e. change order of facets in next step)
genus10.df$type <- factor(genus10.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "J", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
genus10.df <- genus10.df[order(genus10.df$age, genus10.df$age_label), ]
genus10.df$age_label <- factor(genus10.df$age_label, levels = unique(genus10.df$age_label))

# plot
genus_plot <- ggplot(genus10.df, aes(x = age_label, y = Abundance, fill = Genus)) +
  theme_light() +
  geom_col() +
  ggtitle("Genus") +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=9, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=10),
        legend.title = element_text(size=12),
        legend.position = "bottom") +
    scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
genus_plot
```


# Species

* Read RDS file
```{r}
species <- readRDS("species_phyloseq.rds")
```

* Filter for top 10 taxa (excluding 'unclassified')
```{r}
TopNOTUs.Species <- names(sort(taxa_sums(species), TRUE)[2:11])
species10   <- prune_taxa(TopNOTUs.Species, species)
```

## Species Rel. Abundance
```{r}
# melt into df
species10.df <- psmelt(species10)

# remove underscores between words in Species column
species10.df$Species <- gsub("_", " ", species10.df$Species)

# make region a factor in the df (i.e. change order of facets in next step)
species10.df$type <- factor(species10.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "J", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
species10.df <- species10.df[order(species10.df$age, species10.df$age_label), ]
species10.df$age_label <- factor(species10.df$age_label, levels = unique(species10.df$age_label))

# plot
species_plot <- ggplot(species10.df, aes(x = age_label, y = Abundance, fill = Species)) +
  theme_light() +
  geom_col() +
  ggtitle("Species") +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=9, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=9),
        legend.title = element_text(size=12),
        legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3)) +
  scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
species_plot
```


# Segatella plot

* Filter so only Segatella sp are included
```{r}
# Filter phylum level phyloseq object to retain only Segatella species
species.seg <- subset_taxa(species, Genus=="Segatella")

plot_bar(species.seg, fill="Species")

#Remove "empty" samples (<0.1% relative abundance)
species.seg <- prune_samples(sample_sums(species.seg) > 0.1, species.seg)
```

```{r}
# melt into df
species.seg.df <- psmelt(species.seg)

# Filter taxa which have an overall lower relative abundance than 0.1%
species.seg.df <- filter(species.seg.df, Abundance > 0.1)

# remove underscores between words in Species column
species.seg.df$Species <- gsub("_", " ", species.seg.df$Species)

# make region a factor in the df (i.e. change order of facets in next step)
species.seg.df$type <- factor(species.seg.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "J", I = "I", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
species.seg.df <- species.seg.df[order(species.seg.df$age, species.seg.df$age_label), ]
species.seg.df$age_label <- factor(species.seg.df$age_label, levels = unique(species.seg.df$age_label))

# Order the 'Species' factor levels so that UNKNOWN is last
species.seg.df$Species <- factor(species.seg.df$Species, levels = c(sort(unique(species.seg.df$Species[species.seg.df$Species != 'UNKNOWN'])), 'UNKNOWN'))

# find number of unique species
unique_counts_seg <- sapply(species.seg.df, n_distinct)
unique_counts_seg

# plot
seg.species_plot <- ggplot(species.seg.df, aes(x = age_label, y = Abundance, fill = Species)) +
  theme_light() +
  geom_col() +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=7, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 10, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=9),
        legend.title = element_text(size=12),
        legend.position = "right") +
  scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
seg.species_plot
```


# Export Plots - Bacteria

* Save as 300 dpi tiff
```{r}
tiff(file.path(folder_path_plots, "Kingdom_RelAb_290724.tiff"), units="in", width=8.5, height=4, res=300)
plot(king_plot)
dev.off()
```

* Phylum/Genus/Species Patchwork
```{r}
Bact_Tax <- phy_plot / genus_plot / species_plot + plot_layout(heights = c(10, 10, 10)) +
  plot_annotation(tag_levels = 'A')
Bact_Tax
```

* Export Plot
```{r}
tiff(file.path(folder_path_plots, "Bacteria_Top10Taxa.tiff"), units="in", width=9, height=12, res=300)
plot(Bact_Tax)
dev.off()
```

* Export Segatella Plot
```{r}
tiff(file.path(folder_path_plots, "Segatella_TopTaxa.tiff"), units="in", width=8.5, height=4, res=300)
plot(seg.species_plot)
dev.off()
```


# Archaea

## Phylum

```{r}
# Filter phylum level phyloseq object to keep Archaea only
phy.arch <- subset_taxa(phylum, Kingdom=="Archaea")

# prune "empty" samples (0.01%)
phy.arch <- prune_samples(sample_sums(phy.arch) > 0.01, phy.arch)
```

## Faceted bar chart - Phyla
```{r}
# melt into df
phy.arch.df <- psmelt(phy.arch)

# remove underscores between words in Phylum column
phy.arch.df$Phylum <- gsub("_", " ", phy.arch.df$Phylum)

# make region a factor in the df (i.e. change order of facets in next step)
phy.arch.df$type <- factor(phy.arch.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "Jejunum", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
phy.arch.df <- phy.arch.df[order(phy.arch.df$age, phy.arch.df$age_label), ]
phy.arch.df$age_label <- factor(phy.arch.df$age_label, levels = unique(phy.arch.df$age_label))

# plot
phy_plot.arch <- ggplot(phy.arch.df, aes(x = age_label, y = Abundance, fill = Phylum)) +
  theme_light() +
  geom_col() +
  ggtitle("Phylum") +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=9, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=10),
        legend.title = element_text(size=12),
        legend.position = "bottom") +
  scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
phy_plot.arch
```


# Class
```{r}
# Filter class level phyloseq object to keep Archaea only
class.arch <- subset_taxa(class, Kingdom=="Archaea")

# prune "empty" samples (0.01%)
class.arch <- prune_samples(sample_sums(class.arch) > 0.01, class.arch)
```

## Faceted Bar Chart - Class
```{r}
# Melt into df
class.arch.df <- psmelt(class.arch)

# Remove underscores between words in Class column
class.arch.df$Class <- gsub("_", " ", class.arch.df$Class)

# Make region a factor in the df (i.e. change order of facets in next step)
class.arch.df$type <- factor(class.arch.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "Jejunum", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
class.arch.df <- class.arch.df[order(class.arch.df$age, class.arch.df$age_label), ]
class.arch.df$age_label <- factor(class.arch.df$age_label, levels = unique(class.arch.df$age_label))

# Plot
class_plot.arch <- ggplot(class.arch.df, aes(x = age_label, y = Abundance, fill = Class)) +
  theme_light() +
  geom_col() +
  ggtitle("Class") +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=9, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=10),
        legend.title = element_text(size=12),
        legend.position = "bottom") +
  scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
class_plot.arch
```


# Order
```{r}
# Filter order level phyloseq object to keep Archaea only
order.arch <- subset_taxa(order, Kingdom=="Archaea")

# Prune "empty" samples (0.01%)
order.arch <- prune_samples(sample_sums(order.arch) > 0.01, order.arch)
```

## Faceted Bar Chart - Order
```{r}
# Melt into df
order.arch.df <- psmelt(order.arch)

# Remove underscores between words in Order column
order.arch.df$Order <- gsub("_", " ", order.arch.df$Order)

# Make region a factor in the df (i.e. change order of facets in next step)
order.arch.df$type <- factor(order.arch.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "J", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
order.arch.df <- order.arch.df[order(order.arch.df$age, order.arch.df$age_label), ]
order.arch.df$age_label <- factor(order.arch.df$age_label, levels = unique(order.arch.df$age_label))

# Plot
order_plot.arch <- ggplot(order.arch.df, aes(x = age_label, y = Abundance, fill = Order)) +
  theme_light() +
  geom_col() +
  ggtitle("Order") +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=9, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=10),
        legend.title = element_text(size=12),
        legend.position = "bottom") +
  scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
order_plot.arch
```

# Family 

```{r}
# Filter family level phyloseq object to keep Archaea only
family.arch <- subset_taxa(family, Kingdom=="Archaea")

# Prune "empty" samples (0.01%)
family.arch <- prune_samples(sample_sums(family.arch) > 0.01, family.arch)
```

## Faceted Bar Chart - Family
```{r}
# Melt into df
family.arch.df <- psmelt(family.arch)

# Remove underscores between words in Family column
family.arch.df$Family <- gsub("_", " ", family.arch.df$Family)

# Make region a factor in the df (i.e. change order of facets in next step)
family.arch.df$type <- factor(family.arch.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "J", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
family.arch.df <- family.arch.df[order(family.arch.df$age, family.arch.df$age_label), ]
family.arch.df$age_label <- factor(family.arch.df$age_label, levels = unique(family.arch.df$age_label))

# Plot
family_plot.arch <- ggplot(family.arch.df, aes(x = age_label, y = Abundance, fill = Family)) +
  theme_light() +
  geom_col() +
  ggtitle("Family") +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=9, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=10),
        legend.title = element_text(size=12),
        legend.position = "bottom") +
  scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
family_plot.arch
```

# Genus
```{r}
# Filter genus level phyloseq object to keep Archaea only
gen.arch <- subset_taxa(genus, Kingdom=="Archaea")

# prune "empty" samples (0.01%)
gen.arch <- prune_samples(sample_sums(gen.arch) > 0.01, gen.arch)
```

## Faceted bar chart - Genus
```{r}
# Melt into df
gen.arch.df <- psmelt(gen.arch)

# Remove underscores between words in Genus column
gen.arch.df$Genus <- gsub("_", " ", gen.arch.df$Genus)

# Make region a factor in the df (i.e. change order of facets in next step)
gen.arch.df$type <- factor(gen.arch.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "J", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
gen.arch.df <- gen.arch.df[order(gen.arch.df$age, gen.arch.df$age_label), ]
gen.arch.df$age_label <- factor(gen.arch.df$age_label, levels = unique(gen.arch.df$age_label))

# Plot
gen_plot.arch <- ggplot(gen.arch.df, aes(x = age_label, y = Abundance, fill = Genus)) +
  theme_light() +
  geom_col() +
  ggtitle("Genus") +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=9, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=10),
        legend.title = element_text(size=12),
        legend.position = "bottom") +
  scale_fill_manual(values = cb_palette) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
gen_plot.arch
```


# Species
```{r}
# Filter species level phyloseq object to keep Archaea only
species.arch <- subset_taxa(species, Kingdom=="Archaea")

# Prune "empty" samples (0.01%)
species.arch <- prune_samples(sample_sums(species.arch) > 0.01, species.arch)
```

## Faceted Bar Chart - Species
```{r}
# Melt into df
species.arch.df <- psmelt(species.arch)

# Remove underscores between words in Species column
species.arch.df$Species <- gsub("_", " ", species.arch.df$Species)

# Make region a factor in the df (i.e. change order of facets in next step)
species.arch.df$type <- factor(species.arch.df$type, levels=c("D", "J", "I", "C", "PC", "DC"))

# Label function
labels <- c(D = "D", J = "J", I = "Ileum", C = "Caecum", PC = "Proximal Colon", DC = "Distal Colon")

# Order ids by age (using a unique identifier to account for duplicates in df)
species.arch.df <- species.arch.df[order(species.arch.df$age, species.arch.df$age_label), ]
species.arch.df$age_label <- factor(species.arch.df$age_label, levels = unique(species.arch.df$age_label))

# Plot
species_plot.arch <- ggplot(species.arch.df, aes(x = age_label, y = Abundance, fill = Species)) +
  theme_light() +
  geom_col() +
  ggtitle("Species") +
  facet_grid(~ type, space ="free", scales = "free", labeller = labeller(type = labels)) +
  theme(axis.text.x = element_text(angle=90, size=9, hjust=0.95, vjust=0.2), 
        axis.text.y = element_text(size=10),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        plot.title = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        strip.background =element_rect(fill="grey"),
        strip.text = element_text(size = 12, face = "bold"),  # Adjust the size of facet labels
        legend.text.align = 0, 
        legend.text = element_text(size=10),
        legend.title = element_text(size=12),
        legend.position = "bottom") +
  scale_fill_manual(values = cb_palette) +
  guides(fill = guide_legend(nrow = 3)) +
  xlab("Sample ID") + ylab("Relative Abundance (%)")
species_plot.arch
```

# Export Plots - Archaea

* Phylum/Genus/Species Patchwork
```{r}
Arch_Tax <- phy_plot.arch / gen_plot.arch / species_plot.arch + plot_layout(heights = c(10, 10, 10)) +
  plot_annotation(tag_levels = 'A')
Arch_Tax
```

* Export Plot
```{r}
tiff(file.path(folder_path_plots, "Archaea_Top10Taxa.tiff"), units="in", width=9, height=12, res=300)
plot(Arch_Tax)
dev.off()
```
